---
title: "Lab 4: Point clouds from photogrammetry and lidar"
output:
  html_notebook: default
  number_sections: true
author: Hans Ole Ørka, 19. September 2026
editor_options: 
  chunk_output_type: inline
---
<hr style="border:3px solid gray">


<div style="padding: 15px; border: 1px solid transparent; border-color: transparent; margin-bottom: 20px; border-radius: 4px; color: #31708f; background-color: #d9edf7; border-color: #bce8f1;">
**Goal:** Understand difference between point cloud data obtain from lidar or Airborne laser scanning (ALS) and aerial imagery/photogrammetry or Digital Aerial Photogrammetry (DAP). Learn the basic content of the LAS/LAZ format specification and how this data can be manipulated to create Digital Surface Models (DSMs), Digital Terrain Models (DTMs), and Canopy Height Models (CHMs).
</div>


**Data:** The data for the lab can be downloaded from the link below.

- https://eduumb.sharepoint.com/:f:/s/team_mina305instructors/IgD5sXOG76LaTr5QUe5z6ZIhAYwWuhzsMaKkdRar_iUmWd4?e=ya2V67 

# _**Part 1: Exploring point clouds using lidR**_

The lab exercise is carried out using [R](https://www.r-project.org/) and the [lidR-package](https://github.com/r-lidar/lidR). The [lidRBook](https://r-lidar.github.io/lidRbook/) is a fantastic resource! 

## Necessary R packages

```{r,eval=T}
# Load packages 
library("lidR")
library("sf")
library("terra")
library("ggplot2")
library("future")
```

## Read and inspect laser data

Three different point clouds are provided in the folder "data/frydenhaug/":

| Filename    | Type  | p/m^2 |   |   |
|:-------------|-------:|-------:|---|---|
| als2008.laz | lidar | 1.9   |   |   |
| als2021.laz | lidar | 5     |   |   |
| dap2021.laz | image | 20    |   |   |

### ALS 2008

```{r setup, warning=FALSE}
#Read ALS file 
lasFile <- "data/frydenhaug/als2008.laz" # Filename
las <- readLAS(lasFile)
plot(las)

filter <- "-keep_xy 599850 6615750 600250 6616030" #Filter to only select a spesific area  
las <- readLAS(lasFile,filter=filter)
plot(las)
head(las)
names(las)
hist(las$Z)
plot(las,color="Classification")
table(las$Classification)
plot(las,color="PointSourceID")
table(las$PointSourceID)
```
### ALS 2021

```{r}
lasFile <- "data/frydenhaug/als2021.laz" # Filename
filter <- "-keep_xy 599850 6615750 600250 6616030" #Filter to only select a spesific area  
las <- readLAS(lasFile,filter=filter)
plot(las)
head(las)

hist(las$Z)
plot(las,color="Intensity",pal=gray.colors(20))
plot(las,color="Classification")
table(las$Classification)
plot(las,color="PointSourceID")
plot(las,color="ScanAngleRank")
table(las$PointSourceID)
```
### DAP 2021

```{r}
lasFile <- "data/frydenhaug/dap2021.laz" # Filename
filter <- "-keep_xy 599850 6615750 600250 6616030" #Filter to only select a spesific area  
las <- readLAS(lasFile,filter=filter)
las
plot(las)
plot(las,color="RGB")
lasNIR <- las
lasNIR$B <- las$G
lasNIR$G <- las$R
lasNIR$R <- las$Intensity
plot(lasNIR,color="RGB")
table(lasNIR$Classification)
```

# _**Part 2: Create DSM, DTM and CHM**_


```{r}
lasFile <- "data/frydenhaug/als2021.laz" # Filename
filter <- "-keep_xy 599850 6615750 600250 6616030" #Filter to only select a spesific area  
las <- readLAS(lasFile,filter=filter)
```

## Ground classification and creation of digital terrin model
```{r}
# Select ground echoes
ground <- readLAS(lasFile,filter=paste0(filter, " -keep_class 2"))
plot(ground)
x <- plot(ground,color="Intensity",colorPalette=gray.colors(20))
veg <- readLAS(lasFile,filter=paste0(filter, " -drop_class 2"))
plot(veg,add=x)

# DTM
dtm <- rasterize_terrain(las, res = 1, algorithm = tin())
x <- plot(veg)
add_dtm3d(x, dtm)
plot(dtm)

# DSM
dsm <- rasterize_canopy(las,res = 1, algorithm = dsmtin())
plot(dsm)

chm <- dsm - dtm
plot(chm)
```

## Normalizing ALS point cloud data

```{r}
lasdz <- normalize_height(las, tin())
plot(lasdz)
table(lasdz$Classification)
plot(lasdz,color="Classification")
```

## Difference between 2021 and 2008 


```{r}
chm2021 <- chm
lasFile <- "data/frydenhaug/als2008.laz" # Filename
filter <- "-keep_xy 599850 6615750 600250 6616030" #Filter to only select a spesific area  
las <- readLAS(lasFile,filter=filter)
dtm <- rasterize_terrain(las, res = 1, algorithm = tin())
dsm <- rasterize_canopy(las,res = 1, algorithm = dsmtin())
chm2008 <- dsm - dtm
diff <- chm2021 - chm2008
plot(diff)
```


```{r}
chm2021 <- chm
lasFile <- "data/frydenhaug/dap2021.laz" # Filename
filter <- "-keep_xy 599850 6615750 600250 6616030" #Filter to only select a spesific area  
dap <- readLAS(lasFile,filter=filter)
dsmdap <- rasterize_canopy(dap,res = 1, algorithm = dsmtin())
chmdap <- dsmdap - dtm
diff <- chm2021 - chmdap
plot(diff)

```

# **Extra:Repeat on a larger area**


```{r}
library(lidR)
lasFile <- "data/vaaler/0008.las"
filter <- "-keep_xy 606200 6600150 606500 6600450" 
las <- readLAS(lasFile,filter=filter)
plot(las)
head(las)
plot(las,color="Classification")
plot(las,color="PointSourceID")
table(las$Classification)
table(las$ReturnNumber)
plot(las,color="Intensity",pal=gray.colors(20))
```

## Normalizing ALS data

```{r}
lasdz <- normalize_height(las, tin())
plot(lasdz)
table(lasdz$Classification)
plot(lasdz,color="Classification")
lasdz_no_noise <- classify_noise(lasdz,algorithm=ivf(5,2))
table(lasdz_no_noise$Classification)
lasdz_filtered <- filter_poi(lasdz_no_noise,Classification != 7 & Classification != 18)
plot(lasdz_filtered,color="Classification")
plot(lasdz_filtered)
```


## Normalize ALS data from multiple files

This take some time ........

```{r,eval=FALSE}
ctg <- readLAScatalog("data/vaaler/")
opt_filter(ctg) <- "-keep_classification 0 1 2"
opt_output_files(ctg) <-  paste0("data/vaalerdz/", "/{ORIGINALFILENAME}_norm")
ctgdz <- normalize_height(ctg, tin())
```

```{r,echo=FALSE,include=FALSE}
knitr::purl("04_PointClouds.Rmd")
```